---
title: "Traffic Fatality Choropleth"
author: "Mike Silva"
date: "November 16, 2015"
output: 
  html_document: 
    keep_md: yes
---

## Introduction
I wanted to create some choropleth maps of traffic fatality data.  I pulled the latest data from <http://www.iihs.org/iihs/topics/t/general-statistics/fatalityfacts/state-by-state-overview/2013>.  I will create these maps using R.

## Data Management
The first step was to read and summarize in the data:

```{r, message=FALSE}
library(dplyr)
# Read in the HTS data
hts.fatalities.2013 <- read.csv('hts_fatalities_2013.csv', nrows = 1)
hts.fatalities.2013 <- read.csv('hts_fatalities_2013.csv', colClasses = rep('character', length(hts.fatalities.2013))) %>%
  filter(State != 'U.S. total') %>%
  mutate(Deaths.per.100.000.population = as.numeric(Deaths.per.100.000.population)) %>%
  mutate(Deaths.per.100.million.vehicle.miles.traveled = as.numeric(Deaths.per.100.million.vehicle.miles.traveled)) %>%
  mutate(region = tolower(State))
# Generate summaries
summary(hts.fatalities.2013$Deaths.per.100.000.population)
summary(hts.fatalities.2013$Deaths.per.100.million.vehicle.miles.traveled)
```

Deaths per 100 million vehicle miles traveled is roughly one-tenth the size of the deaths per 100,000 million population.  I want to display this data plus quartiles so I need to add quartiles.

```{r}
hts.fatalities.2013$miles.quartile <- with(hts.fatalities.2013, cut(Deaths.per.100.million.vehicle.miles.traveled,  breaks=quantile(Deaths.per.100.million.vehicle.miles.traveled), include.lowest=TRUE))

hts.fatalities.2013$pop.quartile <- with(hts.fatalities.2013, cut(Deaths.per.100.000.population,  breaks=quantile(Deaths.per.100.000.population), include.lowest=TRUE))
```

## Visualizations
```{r, echo=FALSE}
library(ggplot2)

states.map <-map_data('state')

plot1 <- ggplot(hts.fatalities.2013, aes(map_id = region)) +
  geom_map(aes(fill = Deaths.per.100.000.population), map = states.map, color ='black') +
  expand_limits(x = states.map$long, y = states.map$lat) +
  scale_fill_gradient(low='white', high='orange', name='Deaths') +
  guides(fill = guide_colorbar(barwidth = .5, barheight = 10)) +
  theme(legend.position = 'none',
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        axis.text =  element_blank())
  
plot2 <- ggplot(hts.fatalities.2013, aes(map_id = region)) +
  geom_map(aes(fill = Deaths.per.100.million.vehicle.miles.traveled), map = states.map, color ='black') +
  expand_limits(x = states.map$long, y = states.map$lat) +
  scale_fill_gradient(low='white', high='orange', name='Deaths') +
  theme(legend.position = 'none',
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        axis.text =  element_blank())

plot3 <- ggplot(hts.fatalities.2013, aes(map_id = region)) +
  geom_map(aes(fill = as.numeric(pop.quartile)), map = states.map, color ='black') +
  expand_limits(x = states.map$long, y = states.map$lat) +
  scale_fill_gradient(low='white', high='darkgreen', name='Deaths') +
  theme(legend.position = 'none',
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        axis.text =  element_blank())
  
plot4 <- ggplot(hts.fatalities.2013, aes(map_id = region)) +
  geom_map(aes(fill = as.numeric(miles.quartile)), map = states.map, color ='black') +
  expand_limits(x = states.map$long, y = states.map$lat) +
  scale_fill_gradient(low='white', high='darkgreen', name='Deaths') +
  guides(fill = guide_colorbar(barwidth = .5, barheight = 10)) + 
  theme(legend.position = 'none',
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        axis.text =  element_blank())
```

### Deaths per 100,000  

```{r, echo=FALSE}
plot1
```

### Deaths per 100 Million Miles  

```{r, echo=FALSE}
plot2
```

### Deaths per 100,000 Quartiles  

```{r, echo=FALSE}
plot3
```

### Deaths per 100 Million Miles Quartiles  

```{r, echo=FALSE}
plot4
```

### All Together  

```{r, echo=FALSE, fig.align='center',fig.width=12,fig.height=10}
library(gridExtra)
plot1 <- plot1 + ggtitle('Deaths per 100,000')
plot2 <- plot2 + ggtitle('Deaths per 100 Million Miles')
plot3 <- plot3 + ggtitle('Deaths per 100,000 Quartiles')
plot4 <- plot4 + ggtitle('Deaths per 100 Million Miles Quartiles')
grid.arrange(plot1, plot2, plot3, plot4, ncol=2)
```

